\documentclass[11pt]{article}
\usepackage{amsmath,amssymb,amsthm,amsfonts}
\usepackage{geometry}
\usepackage{enumitem}
\usepackage{setspace}
\geometry{margin=0.75in}
\onehalfspacing

\begin{document}

\title{Optimization in Industry \\Mid Semester Exam}
\author{Utpalraj Kemprai\\
MDS202352}
\date{}
\maketitle

\section*{Question 1}

We will introduce a binary variable \(I \in \{0,1\}\). And set the following linear constraint.
\[
    x = 15 + 5\times I
\]

This will ensure that \(x\) is either 15 or 20.

\section*{Question 2}

\subsection*{(A) A and B are parallel}
\begin{itemize}
    \item Net Lead Time = max(LT\_A, LT\_B)
    \item Net Capacity = Cap\_A + Cap\_B
\end{itemize}
\subsubsection*{Assumption for the parallel case}
Both processes A and B must be completed before operation X is considered complete.

\subsection*{(B) A and B are in series}
\begin{itemize}
    \item Net Lead Time = LT\_A + LT\_B
    \item Net Capacity = min(Cap\_A, Cap\_B)
\end{itemize}

% \newpage
\section*{Question 3}
As X and Y are integers, 2X + 2Y is an even integer and so cannot be equal to 35. Since 2X+2Y is always even, we can tighten the constraint by replacing 35 with 34 in the original constraint.

So the new tighter constraint is,
\begin{align*}
    2X + 2Y \leq 34\\
    \text{ or } X + Y \leq 17
\end{align*}

\section*{Question 4}
\subsection*{Sets}
\begin{itemize}
    \item $I$: set of items indexed by \(i\).
    \item $D$: set of days indexed by $d$, where d ranges from the item's availDate to its shipDate and has been coded as an integer (i.e., $d = 1,2,\cdots$). Here 1 corresponds to the earliest date in D, 2 to the day immediately after and so on. D contains all the dates from the earliest availDate (of the items) to the latest shipDate (of the items).
    \item $T$: set of Trucks available indexed by \(t\).
\end{itemize}

\subsection*{Parameters}
\begin{itemize}
    \item \(w_i > 0\): weight of item \(i\).
    \item \(\text{avail}_i\): earliest possible shipping day for item i.
    \item \(\text{ship}_i\): latest shipping day for item i.
    \item Truck load bounds: \(L_{\text{min},t}\) and \(L_{\text{max},t}\), minimum and maximum truck load for truck \(t\).
\end{itemize}

\subsection*{Decision Variables}
\begin{itemize}
    \item \(x_{i,d,t} \in \{0,1\}\): 1 if item \(i\) is shipped on day \(d\) by truck \(t\), otherwise 0.
    \item \(y_{t,d} \in \{0,1\}\): 1 if the truck \(t\) is dispatched on day \(d\), otherwise 0.
\end{itemize}

\subsection*{Constraints}
\begin{enumerate}
    \item \textbf{Item Assignment:}
    \[
        \sum_{d=\text{avail}_i}^{\text{ship}_i}\sum_{t \in T} x_{i,d,t} = 1, \quad \forall i \in I
    \]

    \[
        x_{i,d,t} = 0, \quad d \notin \{\text{avail}_i,\text{avail}_i + 1, \cdots, \text{ship}_i\},\forall i \in I, \forall t \in T
    \]

    \noindent\textbf{Explanation}  

    The above constraints ensure that each item \(i\) will be shipped in one of the dates from \(\text{avail}_i\) to \(\text{ship}_i\).
    
    \item \textbf{Truck Capacity (if the truck is dispatched):}
    \[
        \sum_{i \in I} w_i\cdot x_{i,d,t} \leq L_{\text{max},t} \cdot y_{t,d}, \quad \forall d \in D, \forall t \in T
    \]

    \[
        \sum_{i \in I} w_i\cdot x_{i,d,t} \geq L_{\text{min},t} \cdot y_{t,d}, \quad \forall d \in D, \forall t \in T
    \]

    \noindent\textbf{Explanation}
    
    The above constraints ensure that for each scheduled dispatch for a truck \(t\), the items assigned exceed the minimum load but do not exceed its maximum load. It also ensures that no items are assigned to a truck if the truck is not dispatched.

\end{enumerate}

\subsection*{Objective}
Minimize the cumulative time between the date of shipment of items and their availDate:
\[
    \min \sum_{d \in D} \sum_{i \in I} \sum_{t \in T} (d - \text{avail}_i) \cdot x_{i,d,t}
\]


\section*{Question 5}
\subsection*{Possible Constraints}
\begin{itemize}
    \item \textbf{Limited aircrafts:} Total aircrafts assigned \(\leq\) number of available aircrafts.
    \item \textbf{Flight coverage:} Each scheduled flight must be assigned an aircraft.
    \item \textbf{Meeting demand:} Matching aircraft type (90-seat or 180-seat) with flight demand.
    \item \textbf{Maintenance and crew:} Scheduling constraint for maintenance and crew availability. e.g. time between consecutive flights, number of flights in a day etc.
    \item \textbf{Operational Regulations:} Compliance with regulatory requirements.
    \item \textbf{Boarding and Unboarding Constraints:} Time window for the passengers to board and unboard the flights.
\end{itemize}

\subsection*{Possible Objectives}
\begin{itemize}
    \item \textbf{Maximize Revenue/Profit:} By optimally matching aircraft to flight demand.
    \item \textbf{Minimize Costs:} Including fuel, crew and maintenance costs.
    \item \textbf{Enhance Robustness:} Mitigate delays and disruptions while maintaining the schedule.
\end{itemize}
% end of exam

\end{document}




